---
title: "demo for week 3 (or 4?)"
author: "timothy leffel  &nbsp;&nbsp;&nbsp;&nbsp; may20/2017"
date: ""
output: html_document
css: ../css/notes.css
---

<hr style="height:2px; background-color: gray; color:gray;">

<!-- for some reason this causes problems! -->
<!-- <link href="https://afeld.github.io/emoji-css/emoji.css" rel="stylesheet"> -->

### a mystery

mystery: what does this (mysterious) code do?! ([source](https://github.com/DataScienceSpecialization/courses/blob/master/07_RegressionModels/02_04_residuals_variation_diagnostics/index.Rmd#example-described-by-stefanski-tas-2007-vol-61))

```{r eval=FALSE}
dat <- read.table('http://www4.stat.ncsu.edu/~stefanski/NSF_Supported/Hidden_Images/orly_owl_files/orly_owl_Lin_4p_5_flat.txt', header = FALSE)
fit <- lm(V1 ~ . - 1, data = dat); plot(predict(fit), resid(fit), pch = '.')
```

it works fine and is very compact, but imo it's pretty hard to read. also, the output (*which is a mystery* doesn't convey the point of why the code is interesting, and it just isn't as pretty as it could be). let's try to solve the mystery by making the code a bit clearer (read: more pedagogical/verbose). along the way, we'll use some packages we're already learning, and see examples of concepts we've been working with. 

first let's get the data from the url given in the first chunk.

```{r}
# get the link as a character string (break it up for legibility)
lnk <- paste0(
  "http://www4.stat.ncsu.edu/~stefanski/",
  "NSF_Supported/Hidden_Images/orly_owl_files/orly_owl_Lin_4p_5_flat.txt"
)
# read in the dataset from the link
dat <- read.table(lnk, header=FALSE, sep="", col.names=c("y", paste0("x", 1:4)))

# fit a model predicting y from the x's (`mob` for "model object")
mob <- lm(y ~ x1 + x2 + x3 + x4, data=dat)

# look at a summary table of the model
knitr::kable(summary(mob)$coef)
```


let's also add columns to dat for mob's fitted values and residuals. 
this lets us compare the actual y-vals to the model's, and to evaluate the fit

```{r}
dat$mob_fitted <- fitted(mob)
dat$mob_resid  <- residuals(mob)
```


let's clean up the table so we can put it next to a diagnostic plot

```{r}
# look at the summary table, save it as `mob_coef`
mob_coef <- data.frame(round(coefficients(summary(mob)), digits=3))

# clean up the summary table: get rownames as a col, rename the cols
mob_coef            <- cbind(row.names(mob_coef), mob_coef)
row.names(mob_coef) <- NULL
names(mob_coef)     <- c("variable", "coefficient", "se", "t stat", "p-value")

# want to have signif stars, so write a function to compute them 
get_stars <- function(p){
  # breaks must be longer than labels by one, "(" is not incl; "]" is incl
  cut(p, breaks=c(-.01,.001,.01,.05,.1,1), labels=c("***","**","*",".","n.s."))
}

# now add significance stars to cleaned up summary table (colname is just " ")
mob_coef$` ` <- get_stars(mob_coef$`p-value`)

# reformat pvals so that vals rounded to zero display as "<.001"
mob_coef$`p-value` <- ifelse(mob_coef$`p-value`==0, "<.001", mob_coef$`p-value`)
```

this is all done with base functions. we could do most or all of this munging in a cleaner and easier to read format, by using `magrittr::`/`dplyr::` pipe-chains. here's a first step to get you started:

```{r eval=FALSE, message=FALSE}
# this code isn't evaluated because we set the chunk option `eval=FALSE`
library("magrittr"); library("dplyr")
# the first few steps...
summary(mob) %>% coefficients() %>% round(digits=3) %>% data.frame()
```

but anyway, back to solving the mystery. now that we have the fitted values and residuals 
as columns of `dat`, we can make the scatterplot built above with base graphics 
using the command `plot(predict(fit), resid(fit), pch = '.')`.

```{r}
# load packages for plotting the results the way we want to
library("ggplot2"); library("grid"); library("gridExtra")

# set the theme for all `ggplot()` objects in this session
theme_set(theme_classic())

# build the plot and assign it to `plob` (for "plot object")
(plob <- ggplot(dat, aes(x=mob_fitted, y=mob_resid)) + 
    geom_point(size=.5, shape=4))
```

waaaaaaaow orly wth is?!?!?! IS A OWL OR IS PIGEON OR WHA?! :o <i class="em em-scream_cat"></i>

now we have the main plot. an important point about p-values! (but more importantly: lolol)

now let's format the summary table so we can save it in a figure, along with the plot. let's also format the model call, so that people can see what the dataset was and what the model specification was.

```{r}
# build a coef table grob and assign it to `coef_grob`
coef_grob <- tableGrob(mob_coef, rows=NULL, theme=ttheme_minimal())

# build a text grob that specifies the model formula [as.expression(mob$call)]
model_grob <- textGrob(as.character(as.expression(mob$call)))
```

now let's arrange the main plot with the coefficient table and the model specification. 
seems to me like the best way would be vertically stacked, with the table on top, the 
model call in the middle, and the plot on the bottom. we do this with the 
`gridExtra::grid.arrange()` function. you'll need to use the trial-and-error method if you apply this 
strategy to your own problem.

```{r, fig.height=7, fig.width=5}
# arrange plob and model_grob and coef_grob to display nicely
(output <- grid.arrange(coef_grob, model_grob, plob, nrow=3, heights=c(2,.2,3)))
```

and there we have it!!!

<!-- **exercise**: look at the output we'll save in the next chunk, and then compare  -->
<!-- it to how the `output` figure gets rendered from the chunk above. why is the  -->
<!-- one above smooshed? how can we make it display like we want it to? (i.e. with  -->
<!-- the dimensions we save it with below) -->

we'll save the output in three different formats: pdf, png, and jpg. we're mostly 
doing this just to illustrate how using the `*apply()` family of functions 
can save you keystrokes for allll kinds of purposes. 

```{r}
# save `output` grob in three diff formats, w vectorized loop via `sapply()`
# [use .pdf when not in .rmd]
sapply(c("pdf","png","jpg"), function(x){
  ggsave(plot=output, filename=paste("orly_plot", x, sep="."), device=x, 
         path="plots/", width=5, height=7, units="in")
})
```

**exercise**: write the `sapply()` call as a regular `for`-loop.

**exercise**: modify the `sapply()` loop so that we save the grob in one format, but in 
several different sizes.

**exercise**: modify the `sapply()` loop so that instead of iterating directly over file formats, 
we're iterating over numbers that index the elements of a vector of file formats.

**exercise**: we could've given `gridExtra::grid.arrange()` any number of unnamed 
arguments at the beginning of the call (i.e. we could've just kept specifying more and more  
grobs after `coef_grob`, `model_grob`, `plob`). sometimes we can do this and sometimes 
we can't -- e.g. compare `mean(1,2,3)` with `mean(c(1,2,3))`; then do the same for 
`sum(1,2,3)` and `sum(c(1,2,3))`. figure out what is going on here. (*hint*: when googling, use the keyword "ellipses")

<br>
<hr>

### credits:

- original concept, dataset, and plot by [Leonard Stefanski](http://www4.stat.ncsu.edu/~stefanski/), 
described [here](http://www4.stat.ncsu.edu/~stefanski/NSF_Supported/Hidden_Images/Residual_Surrealism_TAS_2007.pdf). a collection of other examples is [here](http://www4.stat.ncsu.edu/~stefanski/NSF_Supported/Hidden_Images/stat_res_plots.html)

- R code in first chunk from Johns Hopkins data science specialization materials ([link](https://github.com/DataScienceSpecialization/courses/blob/master/07_RegressionModels/02_04_residuals_variation_diagnostics/index.Rmd#example-described-by-stefanski-tas-2007-vol-61))

<br><br><br><br>




<link rel="stylesheet" type="text/css"
href="https://fonts.googleapis.com/css?family=Open+Sans:300,400,400i,700">

<link href="https://fonts.googleapis.com/css?family=Roboto+Mono:300,400,500" rel="stylesheet">

  <style>
body {
  padding: 10px;
  font-size: 12pt;
  font-family: 'Open Sans', sans-serif;
}

h1 { 
  font-size: 20px;
  color: DarkGreen;
  font-weight: bold;
}

h2 { 
    font-size: 16px;
    color: green;
}

h3 { 
  font-size: 24px;
  color: green;
  font-weight: bold;
}

code {
  font-family: 'Roboto Mono', monospace;
  font-size: 14px;
}

pre {
  font-family: 'Roboto Mono', monospace;
  font-size: 14px;
}

</style>


