---
title: "a skeleton to fill in with your own content!"
author: "your name"
output: html_document
---

<!-- notice that comments are in html format here.** 
     if the output type was 'pdf_document', we'd be using 
     LaTeX style for comments (i.e. start line with %) -->

## 1. setting things up

### 1.1 set global chunk options and load packages

<!-- Here's a setup chunk, which won't appear in the output because it has the chunk option `include=FALSE`.  -->

```{r setup, include=FALSE}
# **but comments inside a code chunk are normal R style

# echo=TRUE: the code you type in a chunk will appear in the output
knitr::opts_chunk$set(echo=TRUE) 

# another useful one when you're learning:
# tidy=TRUE: auto-formats code in output so that it's nicely indented etc.
#            (note: this will give you R Studio's preferred code format)


# notice that this chunk has the option include=FALSE
# this means that it won't show up in the output 
# (since it's not really "content", but just controls output)
```

<!-- Now we load a few packages that we'll use below. We specify `message=FALSE` and `warning=FALSE` on this one so that we don't get the annoying package startup messages: -->

```{r attach_packages, message=FALSE, warning=FALSE}
library("dplyr"); library("reshape2"); library("ggplot2"); library("knitr")

# also set the ggplot2 theme to something nicer than the default
theme_set(theme_minimal(base_size=20))
```


### 1.2 read in dataset and inspect

<!-- read in a dataset and print the head here -->

```{r}
# read in a dataset

# print the first few rows
```


### 1.3 clean up dataset as desired

<!-- if there's nothing obvious to clean up, maybe just change some of the column names -->

```{r}
# do some cleaning operations here -- e.g. 
#   - change column names;
#   - recode a column;
#   - reshape the data to long-format with melt(); 
#   - convert factor columns to character or vice versa; 
#   - ...
```



## 2. summarize and visualize columns of interest


### 2.1 display some tables and summaries

<!-- this is  a good place for a couple of `table()` calls -->

```{r}
# make a table by calling `table()` on a column of interest
```


```{r}
# add some stuff to the `table` command above to make a formatted version:
#   - wrap it in `as.data.frame()`, and then wrap that in `kable()`
#   - then give `kable()` a vector of `col.names=` that you want displayed
#   - you can also specify `digits=` in `kable()`, for rounding in the output
```


<!-- you might want to make some group-wise summaries with `aggregate()` and then display them with `kable()` here, too -->


```{r}
# example of combining `aggregate()` with `kable()`:
cars_mpg_tab <- aggregate(mpg ~ gear + cyl, data=mtcars, FUN="mean")

# desired column names for output table:
my_colnames <- c("number of gears", "number of cylinders", "gas mileage (mpg)")

# display the table (note kable() comes from the knitr:: package)
kable(cars_mpg_tab, col.names=my_colnames, align=c("c","l","l"), digits=1)
```


### 2.2 display some plots

<!-- note that the chunk options here cause the plots to be displayed side-by-side -->

```{r plotz, fig.width=9, fig.height=6, out.width='48%', fig.show="hold"}
# distribution of frequency by part of speech
ggplot(mtcars, aes(x=factor(cyl), y=mpg)) +
  geom_boxplot() + 
  theme(plot.title=element_text(face="italic")) +
  labs(title="plot title",
       subtitle="subtitle",
       caption="a caption with some info or a statement")

# distribution of rank by part of speech
ggplot(mtcars, aes(x=factor(gear), y=mpg)) +
  geom_boxplot() + 
  theme(axis.text.x=element_text(color="blue")) +
  labs(x="x axis label",
       y="y axis label")
```

<!-- here's the corresponding commands to create these plots with base graphics (this chunk is not evaluated because of the chunk option `eval=FALSE`; but it is displayed because we set `echo=TRUE` in the setup chunk): -->

```{r base_plotz, eval=FALSE}
# same as above plots, but with base graphics
# (not displayed bc of chunk option eval=FALSE)
boxplot(mpg ~ factor(cyl), data=mtcars)
boxplot(mpg ~ factor(gear), data=mtcars)
```



## 3. perform some analysis (optional -- only if it'd be useful for you!)


### 3.1 fit a model to the data 

```{r}
# fit a linear regression model with `lm()` 
# the outcome is before the tilde; and 
# the predictors are after the tilde, separated by "+"

# search the help window and scroll to bottom of `lm()` page to see examples
```

### 3.2 examine the model summary and evaluate relationships

```{r}
# look at the full model summary with `summary(your_model)`
```



### 3.3 evaluate the overall quality of the model fit 


```{r}
# extract the r-squared value from the model summary
# summary(your_model)$r.squared

# or compute it directly (see week1 notes for pearson's r)
# cor(your_data$outcome_var, fitted(your_model))^2
```


## 3.4 summarize the findings

<!-- e.g. with some bullet points -->



